---
title: "PCA"
author: "Angel Allen"
date: "2022-11-15"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, error = FALSE)
```

### Packages

```{r packages, message=F}
# importing the data
library(readxl) # import
library(tidyverse) # import

# visualisation
library(ggplot2) # ggplot()
library(ggpubr) # ggarrange()
library(ggrepel)
library(ggfortify)
library(kableExtra)

# data manipulation
library(rcompanion) # transformTukey()
# library(broom) # augment()
library(psych) #principal()
library(lme4) # lmer()

# library(esquisse)

theme_set(theme_bw()) # set ggplot to b/w
# options(scipen = 100) # set to 0 to go back to scientific notation
# rm(list = ls())
```

```{r import}
source("scripts/import.R")

bad_fish <- c("56.2-6.10", "5.2-6.10", "71.4-7.10", "5.4-7.10")

raw_11 <- read_data("11-Oct-22") %>% subset(.$fish_id %in% grep("6.10", .$fish_id, value = T))

raw_11 <- raw_11[!raw_11$fish_id %in% bad_fish,]

raw_12 <- read_data("12-Oct-22") %>% subset(.$fish_id %in% grep("7.10", .$fish_id, value = T))

raw_12 <- raw_12[!raw_12$fish_id %in% bad_fish,]

raw_all <- bind_rows(raw_11, raw_12, .id = "Family")
```

First I'm going to subset the data from the dark time bins.

```{r dark data}
dark_bins <- c(6:10)
# dark_bins <- c(1:5)
# dark_bins <- c(1:15)

raw_dark <- raw_all %>% subset(Bin %in% dark_bins)
```

# Scores

```{r motor score}
# locomotor variables
# mvars <- colnames(raw_dark)[c(9:18, 26:33)]
variances <- colnames(raw_dark)[grep("var", colnames(raw_dark))]
novar <- colnames(raw_12)[!colnames(raw_12) %in% variances]
mvars <- novar[c(8:11, 13, 14, 20, 21)]

p_mscore <- principal(raw_dark[,mvars])
raw_dark$mscore <- p_mscore$scores %>% as.numeric()
```

I'd like to potentially add a startle response to the anxiety score, which would be the *change* in distance travelled/velocity/other locomotor variables from the light to dark transition.

```{r anxiety score}
# anxiety-related variables
avars <- novar[15:19]

p_ascore <- principal(raw_dark[,avars])
raw_dark$ascore <- p_ascore$scores %>% as.numeric()
```

```{r prcomp scores}
prcomp(raw_dark[,mvars]) %>%
  summary() %>% 
  .$importance %>% 
  as.data.frame() %>% 
  t() %>% 
  as.data.frame() %>% 
  rownames_to_column("PC") %>% 
  mutate(`Proportion of Variance` = `Proportion of Variance`*100, 
         `Cumulative Proportion` = `Cumulative Proportion`*100) %>% 
  ggplot(aes(x = reorder(PC, -`Proportion of Variance`))) +
  geom_col(aes( y = `Proportion of Variance`)) +
  labs(y = "Proportion of Variance (%)", 
       x = "Principal Component") +
  geom_point(aes(y = `Cumulative Proportion`))+
  geom_line(aes(y = `Cumulative Proportion`, group = 1), show.legend = TRUE) +
  theme_bw() +
  theme(aspect.ratio = 1) +
  rotate_x_text(angle = -45)
# scree plots shows that PC1 and 2 descripte ~90% of the variance, so will plot those against one another

prcomp(raw_dark[,mvars]) %>%
  autoplot(data = raw_all,
    colour = "Genotype",
    # shape = "Genotype",
    size = 2) +
  theme_bw() + 
  theme(aspect.ratio = 1) +
  labs(shape = "Genotype")

raw_dark$PC1 <- prcomp(raw_dark[,mvars])$x[,"PC1"]
raw_dark$PC2 <- prcomp(raw_dark[,mvars])$x[,"PC2"]

ggplot(raw_dark, aes(PC1, PC2, colour = Genotype, label = fish_id)) +
  geom_point(size = 2) +
  geom_text_repel()
```

```{r factors}
# corPlot(raw_dark[,mvars])
# vss(raw_dark[,mvars])
# fa(raw_dark[,mvars])
# iclust(raw_dark[,mvars])
# 
# corPlot(raw_dark[,avars])
# vss(raw_dark[,avars])
# fa(raw_dark[,avars])
# iclust(raw_dark[,avars])
```

The factor analysis for the locomotor score shows that up to 90% of the variation in the data can be explained by 1 factor, which is very good. For the anxiety features and score, an "ultra-Heywood case was detected", which apparently invalidates the factor analysis.

```{r plot mscore}
source("scripts/graphs.R")

# vars <- c("Genotype", "mscore", "Trial")

print("Motor score")
boxplot_data(raw_dark, "Genotype", "mscore", "Family")
print("Anxiety score")
boxplot_data(raw_dark, "Genotype", "ascore", "Family")

# ggplot(raw_dark, aes(x = Genotype, y = mscore, fill = Family)) +
#   geom_jitter(aes(colour = Trial), size = 2, alpha = 0.5) +
#   geom_boxplot(alpha = 0.25, outlier.shape = NA) +
#   ggtitle("Motor score")

ggplot(raw_dark) +
  geom_point(aes(x = mscore, y = ascore, colour = Genotype), size = 2)
```

```{r raw anova}
model <- lm(mscore ~ Genotype + Family + Trial + Genotype:Trial + Genotype:Family, data = raw_dark)
summary(model)

aov(model, data = raw_dark) %>% summary()

# model2 <- lmer(mscore ~ Genotype + Family + Trial + Genotype:Trial + Genotype:Family)
```

## Summarised data

```{r mean dark}
source("scripts/DE functions.R")

raw_mean_dark <- raw_dark %>%
  group_by(Family, fish_id, Genotype, Position, Tray, Trial, Time) %>%
  summarise_if(is.numeric, mean) %>% ungroup()

lambdas <- get_lambda(raw_mean_dark, novar[-c(1:7)]) %>% t()

mean_dark <- raw_mean_dark
for (i in novar[-c(1:7)]) {
  lambda <- lambdas[i, 1]
  mean_dark <- mean_dark %>%
    mutate_at(i, function(x){x^lambda})
}

mean_dark
```

```{r mean scores}
# locomotor variables
p_mscore <- principal(mean_dark[,mvars])
mean_dark$mscore <- p_mscore$scores %>% as.numeric()

# anxiety-related variables
p_ascore <- principal(mean_dark[,avars])
mean_dark$ascore <- p_ascore$scores %>% as.numeric()
```

Distribution of the scores

```{r scores distribution}
plot_histbox(mean_dark, "mscore")
plot_histbox(mean_dark, "ascore")

a <- ggplot(mean_dark[mean_dark$Family == 2,]) +
  geom_freqpoly(aes(x = mscore, colour = Genotype), binwidth = 0.4, size = 1) +
  # facet_wrap(~ Family) +
  ggtitle("Motor score")

b <- ggplot(mean_dark[mean_dark$Family == 2,]) +
  geom_freqpoly(aes(x = ascore, colour = Genotype), binwidth = 0.4, size = 1) +
  # facet_wrap(~ Family) +
  ggtitle("Anxiety score")

ggarrange(a, b, common.legend = T)
```

```{r mean prcomp}
prcomp(mean_dark[,mvars]) %>%
  summary() %>% 
  .$importance %>% 
  as.data.frame() %>% 
  t() %>% 
  as.data.frame() %>% 
  rownames_to_column("PC") %>% 
  mutate(`Proportion of Variance` = `Proportion of Variance`*100, 
         `Cumulative Proportion` = `Cumulative Proportion`*100) %>% 
  ggplot(aes(x = reorder(PC, -`Proportion of Variance`))) +
  geom_col(aes( y = `Proportion of Variance`)) +
  labs(y = "Proportion of Variance (%)", 
       x = "Principal Component") +
  geom_point(aes(y = `Cumulative Proportion`))+
  geom_line(aes(y = `Cumulative Proportion`, group = 1), show.legend = TRUE) +
  theme_bw() +
  theme(aspect.ratio = 1) +
  rotate_x_text(angle = -45)
# scree plots shows that PC1 and 2 descripte ~90% of the variance, so will plot those against one another
  
mean_dark$PC1 <- prcomp(mean_dark[,mvars])$x[,"PC1"]
mean_dark$PC2 <- prcomp(mean_dark[,mvars])$x[,"PC2"]

ggplot(mean_dark, aes(PC1, PC2, colour = Genotype, label = fish_id)) +
  geom_point(size = 2) +
  geom_text_repel()
```

```{r mean plots}
ggplot(mean_dark) +
  geom_point(aes(x = mscore, y = ascore, colour = Genotype), size = 2)

vars <- novar[-c(1:7)]

lapply(vars, boxplot_data, data = mean_dark, x = "Genotype", points = "Family")
```

ANOVA using scores

```{r mean anova}
# coding genotype as numeric
model <- lm(mscore ~ Genotype + Trial + Genotype:Trial, data = mean_dark)
summary(model)

aov(model, data = mean_dark) %>% summary()
```
